Structure-Preserving Model-Reduction of Dissipative Hamiltonian Systems

### 2017-03-19 dissipative_hamiltonian/chapters/4.numer.tex
first draft finished

We discretize the torus into $N_{\Delta x}$ equidistant points and define $\Delta x = L/N_{\Delta x}$, $x_i = i\Delta x$, $q_i=q(t,x_i)$ and $p_i=p(t,x_i)$ for $i = 1, \dots, N_{\Delta x}$. The discretization of $r$ corresponds to a diagonal and semi-positive definite matrix $r_\Delta$. Further, we discretize (\ref{eq:4.1}) using a standard centeral finite differences schemes to obtain
\begin{equation} \label{eq:4.2}
	\dot z = \mathbb J_{2n} K^T K z - R z,
\end{equation}
where $z = (q_1,\dots,q_{N_{\Delta x}},p_1,\dots,p_{N_{\Delta x}})$ and $K$ and $R$ are given as
\begin{equation} \label{eq:4.3}
	K^T K =
	\begin{pmatrix}
		I & 0 \\
		0 & c^2D_x^TD_x
	\end{pmatrix} , \quad
	R =
	\begin{pmatrix}
		0 & 0 \\
		0 & r_\Delta
	\end{pmatrix},
\end{equation}
with $D_x^TD_x = D_{xx}$ the central finite differences matrix operator. Writing (\ref{eq:4.2}) in a TDD formulation gives
\begin{equation} \label{eq:4.4}
\begin{aligned}
	& \dot z = \mathbb J_{2n} K^T f(t) \\
	& f(t) + R \int_0^t f(t) = K z.
\end{aligned}
\end{equation}
Here, since $R$ is not time dependent, it commutes with the integration operator. The Hamiltonian extension of (\ref{eq:4.4}), then takes the form (\ref{eq:2.10.a})-(\ref{eq:2.10.c}).

The initial condition used is given by
\begin{equation} \label{eq:4.5}
	q_i(0) = h( 10\times|x_i - \frac{1}{2}| ), \quad p_i = 0, \quad i=1,\dots,N
\end{equation}
where $h(s)$ is the cubic spline function
\begin{equation} \label{eq:4.6}
h(s) = 
\left\{
\begin{aligned}
& 1 - \frac{3}{2}s^2 + \frac{3}{4}s^3, \quad & 0\leq s \leq 1, \\
& \frac{1}{4}(2-s)^3, & 1< s \leq 2, \\
& 0, & s > 2.
\end{aligned}
\right.
\end{equation}

For the numerical time integration of the extended Hamiltonian system, the Str\"omer-Verlet time stepping scheme (\ref{eq:2.3}) is used. Further in each time step, the system of linear equations (\ref{eq:2.4.1}) is solved to recover $z$. System parameters are summarized below.
\vspace{0.5cm}
\begin{center}
\begin{tabular}{|l|l|}
\hline
Domain length & $L = 1$ \\
No. grid points & $N = 500$ \\
Space discretization size & $\Delta x = 0.002$ \\
Time discretization size & $\Delta t = 0.01$ \\
Wave speed & $c^2 = 0.1$ \\
\hline
\end{tabular}
\end{center}
\vspace{0.5cm}

@@

To construct the reduced basis for different methods, we gather the snapshots of (\ref{eq:4.1}). We construct RDH reduced system according to the Algorithm \ref{alg:3.1}. The cotangent lift is used to construct an ortho-symplectic reduced basis $A$ (\ref{eq:3.5}). The performance of the RDH method is then compared to the POD, and the method proposed in \cite{peng2016geometric} referred to as the PSD.

@@

The $L^2$ error between the full systems and the RDH, the PSD and the POD method are presented in Figure \ref{fig:4.1}.c. We notice that symplectic methods provide a more accurate solution with respect to the POD method. In fact, the POD method does not yield a stable reduced system.  Further it is seen that enriching the PSD reduced basis does not yield a significant enhancement in the accuracy of the reduced system. This is because in the PSD method, a non-conservative system is being numerically integrated with a symplectic integrator. This results in an incorrect evolution of energy and eventually, in a qualitatively wrong numerical solution.

On the other hand, We notice that the RDH method with 40 modes provides a significantly more accurate solution compared to the PSD method with 60 modes. The RDH method provides a conservative reduced system where the dissipated energy is trapped in the hidden strings. The conservation of energy is then guaranteed using a symplectic integrator. Due to this reason, we observe remarkable increase in the accuracy by enriching the RDH reduced basis.

@@

\begin{figure}[t]
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{./figs/wave/solution} & 
\includegraphics[width=0.5\textwidth]{./figs/wave/singular} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{./figs/wave/error} & 
\includegraphics[width=0.5\textwidth]{./figs/wave/energy} \\
(c) & (d) \\
\includegraphics[width=0.55\textwidth]{./figs/wave/energy_conserved} &
\includegraphics[width=0.45\textwidth]{./figs/wave/error_homo} \\
(e) & (f) 
\end{tabular}
\caption{(a) The solution to the original dissipative wave equation (\ref{eq:4.1}), (b) The decay of the singular values for the POD, the PSD and the RDH methods, (c) The $L^2$ error for the different methods, (d) Energy preservation for different methods, (e) Energy preservation of the Hamiltonian extension for the original and the reduced system. ``FM'' and ``RM'' stand for the full model and the reduced model respectively. (f) The $L^2$ error between the solution to the reduced system and the full in a near-zero dissipation regime.} \label{fig:4.1}
\end{figure}

@@

Since $J_{2n}$ is full ranked, one can always find a coordinate transformation $ x= T \tilde x$ such that $T^{-1} J_{2n} T^{-T} = \mathbb J_{2n}$. The dissipative Hamiltonian formulation for (\ref{eq:4.7}) takes the form
\begin{equation} \label{eq:4.8}
	\dot {\tilde x} = \mathbb J_{2n} \tilde Q^T\tilde Q \tilde x - \tilde Rx + \tilde u,
\end{equation}
where $\tilde Q = QT$, $\tilde R = T^{-1}RT^{-T}Q^TQ$ and $\tilde u = T^{-1} u$. Note that in this case, since $T$ is orthogonal and $R$ is diagonal, then $\tilde R$ is symmetric and semi-positive definite. The TDD formulation of (\ref{eq:4.8}) takes the form
\begin{equation} \label{eq:4.9}
	\dot{\tilde x} = \mathbb{J}_{2n} \tilde Q^T f(t) + \tilde u, \quad f(t) + \tilde R \int_0^t f(t) = \tilde Q \tilde x.
\end{equation}
The extended Hamiltonian formulation (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) with a quadratic Hamiltonian $H_{\text{ex}}$ can be then carried out according to section \ref{??}. We note that due to the input $\tilde u$, the Hamiltonian $H_{\text{ex}}$ is time dependent. As the matter of fact $\frac{d}{dt} H_{\text{ex}} = \tilde u^T\tilde Q^T \tilde Q \tilde x$. If we define $\overset{\circ}{H}_{\text{ex}} : \mathbb R^{2n}\times \mathcal H^{2n}\times \mathbb R^{2}\to \mathbb R$ as
\begin{equation} \label{eq:4.10}
	\overset{\circ}{H}_{\text{ex}}(\tilde x,\phi,\theta,t,e) = H_{\text{ex}}(\tilde x,\phi,\theta,t) + e, \quad \dot e = - \partial_t H_{\text{ex}},
\end{equation}
it is easily checked that $\frac d {dt} \overset{\circ}{H}_{\text{ex}} =0$ \cite{Hairer:1250576}. However for the time integration of the Hamiltonian system related to $\overset{\circ}{H}_{\text{ex}}$, since the evolution of $\tilde x$, $\phi$ and $\theta$ does not explicitly depend on $e$, we can apply a symplectic integrator directly on (\ref{eq:4.10}). Thus, the passivity of (\ref{eq:4.7}) will be preserved through a symplectic time integration of (\ref{eq:4.9}).

Using an ortho-symplectic reduced basis $A$ The reduced dissipative Hamiltonian method can be applied to (\ref{eq:4.9}) to construct a reduced system of the form (\ref{eq:3.12.a})-(\ref{eq:3.12.c}) together with the extended Hamiltonian $\tilde H_{\text{ex}}$. We immediately notice that $\frac{d}{dt} \tilde H_{\text{ex}} = (A^+ \tilde u)^T A^T \tilde Q^T \tilde Q A y$, meaning that the reduced system is passive. Further the dissipative Hamiltonian structure of the reduced system indicates that the reduced system also carries a port-Hamiltonian structure.

We consider a 100-dimensional ($n=50$) port-Hamiltonian system for the ladder network discussed above. We take $C_i=1$, $L_i = 1$, $R_i=0.2$ for $i=1,\dots,50$, and $R_{51} = 0.4$. The cotangent lift is used to construct a reduced basis and the RDH method is used to construct a reduced system according to Algorithm (\ref{alg:3.1}).

@@

Reduced bases of size $2k = 10$, $2k = 20$ and $2k = 30$ are constructed with the RDH and the moment matching method. Figure \ref{fig:4.3}.c shows the error in the charge in $C_1$ for the two methods. We observe that although the moment matching method is bounded, over long-time integration, the RDH method provides a significantly more accurate solution. This is because in the moment matching method, passivity of the reduced system implies that the energy of the system will be bounded by the input energy. However, there is no guarantee that the dissipation of energy in the reduced system mimics the one of the original system. On the other hand, the RDH method allows correct dissipation of energy through the hidden strings. Also the symplectic time integration in the RDH method guarantees that the total energy will be preserved.

@@

Figure \ref{fig:4.3}.a and Figure \ref{fig:4.3}.b show the average error in the charge and flux of the capacitors and inductors respectively, for these methods. The RDH method provides a significantly better accuracy compared to the moment matching method. This is because the charge of $C_1$ is specified as the output of interest in the moment matching method so it is expected that that method provides low accuracy for computing other outputs. On the other hand, the RDH method not only provides high accuracy in computing the charge for $C_1$ but also it has high accuracy for computing other outputs of interest.

>>>>

We discretize the torus into $N_{\Delta x}$ equidistant points and define $\Delta x = L/N_{\Delta x}$, $x_i = i\Delta x$, $q_i=q(t,x_i)$ and $p_i=p(t,x_i)$ for $i = 1, \dots, N_{\Delta x}$. The discretization of $r$ corresponds to a diagonal and semi-positive definite matrix $r_\Delta$. Further, we discretize (\ref{eq:4.1}) using a standard centeral finite differences schemes to obtain
\begin{equation} \label{eq:4.2}
	\dot z = \mathbb J_{2n} K^T K z - R z,
\end{equation}
where $z = (q_1,\dots,q_{N_{\Delta x}},p_1,\dots,p_{N_{\Delta x}})$ and $K$ and $R$ are given as
\begin{equation} \label{eq:4.3}
	K^T K =
	\begin{pmatrix}
		I & 0 \\
		0 & c^2D_x^TD_x
	\end{pmatrix} , \quad
	R =
	\begin{pmatrix}
		0 & 0 \\
		0 & r_\Delta
	\end{pmatrix},
\end{equation}
with $D_x^TD_x = D_{xx}$ the central finite differences matrix operator. Writing (\ref{eq:4.2}) in a TDD formulation gives
\begin{equation} \label{eq:4.4}
	\dot z = \mathbb J_{2n} K^T f(t), \quad f(t) + R \int_0^t f(t) = K z.
\end{equation}
Here, since $R$ is not time dependent, it commutes with the integration operator. The Hamiltonian extension of (\ref{eq:4.4}), then takes the form (\ref{eq:2.10.a})-(\ref{eq:2.10.c}).

The initial condition used is given by
\begin{equation} \label{eq:4.5}
	q_i(0) = h( 10\times|x_i - \frac{1}{2}| ), \quad p_i = 0, \quad i=1,\dots,N,
\end{equation}
where $h(s)$ is the cubic spline function
\begin{equation} \label{eq:4.6}
h(s) = 
\left\{
\begin{aligned}
& 1 - \frac{3}{2}s^2 + \frac{3}{4}s^3, \quad & 0\leq s \leq 1, \\
& \frac{1}{4}(2-s)^3, & 1< s \leq 2, \\
& 0, & s > 2.
\end{aligned}
\right.
\end{equation}

For the numerical time integration of the extended Hamiltonian system, the Str\"omer-Verlet time stepping scheme (\ref{eq:2.3}) is used. Further in each time step, the system of linear equations (\ref{eq:2.11}) is solved to recover $z$. System parameters are summarized below.
\vspace{0.5cm}
\begin{center}
\begin{tabular}{|l|l|}
\hline
Domain length & $L = 1$ \\
No. grid points & $N = 500$ \\
Space discretization size & $\Delta x = 0.002$ \\
Time discretization size & $\Delta t = 0.002$ \\
Wave speed & $c^2 = 0.1$ \\
\hline
\end{tabular}
\end{center}
\vspace{0.5cm}

@@

To construct the reduced basis for different methods, we gather the snapshots of (\ref{eq:4.1}). We construct the RDH reduced system according to the Algorithm \ref{alg:3.1}. The cotangent lift is used to construct an ortho-symplectic reduced basis $A$ (\ref{eq:3.5}). The performance of the RDH method is then compared to the POD, and the method proposed in \cite{peng2016geometric} referred to as the PSD.

@@

The $L^2$ error between the full systems and the RDH, the PSD and the POD methods are presented in Figure \ref{fig:4.1}.c. We notice that symplectic methods provide a more accurate solution with respect to the POD method. In fact, the POD method does not yield a stable reduced system.  Further it is seen that enriching the PSD reduced basis does not yield a significant enhancement in the accuracy of the reduced system. This is because in the PSD method, a non-conservative system is being numerically integrated with a symplectic integrator. This results in an incorrect evolution of energy and eventually, in a qualitatively wrong numerical solution.

On the other hand, We notice that the RDH method with 40 modes provides a significantly more accurate solution compared to the PSD method with 60 modes. The RDH method provides a conservative reduced system where the dissipated energy is absorbed by the hidden strings. The conservation of energy is then guaranteed using a symplectic integrator. Due to this reason, we observe remarkable increase in the accuracy by enriching the RDH reduced basis.

@@

\begin{figure}[t]
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{./figs/wave/solution} & 
\includegraphics[width=0.5\textwidth]{./figs/wave/singular} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{./figs/wave/error} & 
\includegraphics[width=0.5\textwidth]{./figs/wave/energy} \\
(c) & (d) \\
\includegraphics[width=0.55\textwidth]{./figs/wave/energy_conserved} &
\includegraphics[width=0.45\textwidth]{./figs/wave/error_homo} \\
(e) & (f) 
\end{tabular}
\caption{(a) The solution to the original dissipative wave equation (\ref{eq:4.1}), (b) The decay of the singular values for the POD, the PSD and the RDH methods, (c) The $L^2$ error for the different methods, (d) Energy preservation for different methods, (e) Energy preservation of the Hamiltonian extension for the original and the reduced system. ``FM'' and ``RM'' stand for the full model and the reduced model respectively. (f) The $L^2$ error between the solution to the reduced system and the full system in a near-zero dissipation regime.} \label{fig:4.1}
\end{figure}

@@

Since $J_{2n}$ is full ranked, one can always find a coordinate transformation $ x= T \tilde x$ such that $T^{-1} J_{2n} T^{-T} = \mathbb J_{2n}$. The dissipative Hamiltonian formulation for (\ref{eq:4.7}) takes the form
\begin{equation} \label{eq:4.8}
	\dot {\tilde x} = \mathbb J_{2n} \tilde Q^T\tilde Q \tilde x - \tilde Rx + \tilde u,
\end{equation}
where $\tilde Q = QT$, $\tilde R = T^{-1}RT^{-T}Q^TQ$ and $\tilde u = T^{-1} u$. Note that in this case, since $T$ is orthogonal and $R$ is diagonal, then $\tilde R$ is symmetric and semi-positive definite. The TDD formulation of (\ref{eq:4.8}) takes the form
\begin{equation} \label{eq:4.9}
	\dot{\tilde x} = \mathbb{J}_{2n} \tilde Q^T f(t) + \tilde u, \quad f(t) + \tilde R \int_0^t f(t) = \tilde Q \tilde x.
\end{equation}
The extended Hamiltonian formulation (\ref{eq:2.10.a}-\ref{eq:2.10.c}) with a quadratic Hamiltonian $H_{\text{ex}}$ can be then carried out according to section \ref{sec:2.2}. We note that due to the input $\tilde u$, the Hamiltonian $H_{\text{ex}}$ is time dependent. As the matter of fact $\frac{d}{dt} H_{\text{ex}} = \tilde u^T\tilde Q^T \tilde Q \tilde x$. If we define $\overset{\circ}{H}_{\text{ex}} : \mathbb R^{2n}\times \mathcal H^{2n}\times \mathbb R^{2}\to \mathbb R$ as
\begin{equation} \label{eq:4.10}
	\overset{\circ}{H}_{\text{ex}}(\tilde x,\phi,\theta,t,e) = H_{\text{ex}}(\tilde x,\phi,\theta,t) + e, \quad \dot e = - \partial_t H_{\text{ex}},
\end{equation}
it is easily checked that $\frac d {dt} \overset{\circ}{H}_{\text{ex}} =0$ \cite{Hairer:1250576}. However for the time integration of the Hamiltonian system related to $\overset{\circ}{H}_{\text{ex}}$, since the evolution of $\tilde x$, $\phi$ and $\theta$ does not explicitly depend on $e$, we can apply a symplectic integrator directly on (\ref{eq:4.10}). Thus, the passivity of (\ref{eq:4.7}) will be preserved through a symplectic time integration of (\ref{eq:4.9}).

Using an ortho-symplectic reduced basis $A$ The Reduced Dissipative Hamiltonian method can be applied to (\ref{eq:4.9}) to construct a reduced system of the form (\ref{eq:3.12.a}-\ref{eq:3.12.c}) together with the extended Hamiltonian $\tilde H_{\text{ex}}$. We immediately notice that $\frac{d}{dt} \tilde H_{\text{ex}} = (A^+ \tilde u)^T A^T \tilde Q^T \tilde Q A y$, meaning that the reduced system is passive. Further the dissipative Hamiltonian structure of the reduced system indicates that the reduced system also carries a port-Hamiltonian structure.

We consider a 100-dimensional ($n=50$) port-Hamiltonian system for the ladder network discussed above. We take $C_i=1$, $L_i = 1$, $R_i=0.2$ for $i=1,\dots,50$, and $R_{51} = 0.4$. The cotangent lift is used to construct a reduced basis and the RDH method is used to construct a reduced system according to Algorithm \ref{alg:3.1}.

@@

Reduced bases of size $2k = 10$, $2k = 20$ and $2k = 30$ are constructed with the RDH and the moment matching method. Figure \ref{fig:4.3}.c shows the error in the charge in $C_1$ for the two methods. We observe that although the moment matching method is bounded, over long-time integration, the RDH method provides a significantly more accurate solution. This is because in the moment matching method, the passivity of the reduced system implies that the energy of the system will be bounded by the input energy. However, there is no guarantee that the dissipation of energy in the reduced system mimics the one of the original system. On the other hand, the RDH method allows correct dissipation of energy through the hidden strings. Also the symplectic time integration in the RDH method guarantees that the total energy will be preserved.

@@

Figure \ref{fig:4.3}.a and Figure \ref{fig:4.3}.b show the average error in the charge and flux of the capacitors and inductors respectively. The RDH method provides a significantly better accuracy compared to the moment matching method. This is because the charge of $C_1$ is specified as the output of interest in the moment matching method and so it is expected that that method provides low accuracy for computing other outputs. On the other hand, the RDH method not only provides high accuracy in computing the charge for $C_1$ but also it has high accuracy for computing other outputs of interest.

### 2017-03-19 dissipative_hamiltonian/chapters/5.conc.tex
first draft finished

Here comes the conclusion

>>>>

In this paper we present the Reduced Dissipative Hamiltonian method. The method preserves the symplectic structure of dissipative Hamiltonian systems and guarantees the correct dissipation of energy through time integration. The canonical heat bath, coupled with the full system, constructs a closed system of evolution, therefore a symplectic time integrator can be applied to preserve the system energy and stability over-long time integration. Moreover, the method allows the usage of a generic ortho-symplectic basis for model reduction.

The numerical simulations illustrate that the RDH method can preserve the system energy with significant more accuracy than other methods. Further they show that the hidden strings assure that the dissipation of energy in the reduce system mimics the dissipation of energy in the full system. This ensures that the local error do not accumulate over long-time integration.

All in all, the Reduced Dissipative Hamiltonian method is a promising model reduction technique for dissipative Hamiltonian systems and can save significant computational cost while preserving the symplectic structure and the dissipation structure of the full system.

### 2017-03-19 dissipative_hamiltonian/paper.tex
first draft finished

\begin{abstract}
Reduced basis methods are popular for approximately solving large and complex systems of differential equations. However, conventional methods do not generally preserve conservation laws and symmetries of the full order model. Here, we present The Reduced Dissipative Hamiltonian Method, that preserves the symplectic symmetry of dissipative Hamiltonian systems over the course of model reduction. This method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath. This results in a correct dissipation of energy, preservation of total energy and ultimately in the stability of the solution. Accuracy and stability of the method is illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.

\keywords{First keyword \and Second keyword \and More}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

>>>>

\begin{abstract}
Reduced basis methods are popular for approximately solving large and complex systems of differential equations. However, conventional methods do not generally preserve conservation laws and symmetries of the full order model. Here, we present The Reduced Dissipative Hamiltonian Method, that preserves the symplectic symmetry of dissipative Hamiltonian systems over the course of model reduction. This method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath. This results in a correct dissipation of energy, preservation of total energy and ultimately in the stability of the solution. Accuracy and stability of the method are illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.

\keywords{Model order reduction \and Symplectic model reduction \and The Reduced Dissipative Hamiltonian method}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}
